---
title: Practical Machine Learning Prediction Assignment.
author: "Alexandr Polyakov"
date: "22.09.2015"
output: html_document
---


# Synopsis

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. 

In this project, we use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants who were asked to perform barbell lifts correctly and incorrectly in 5 different ways.

This project aims to analyse data from accelerometers to predict the manner in which the participants did the exercise. 

The data for this project can be obtained from this source: http://groupware.les.inf.puc-rio.br/har. 

# Data Processing

###1. Preliminary settings

```{r}
#make code visible
echo = TRUE
#set up directory for graphs
knitr::opts_chunk$set(fig.path='figures/')
cache = TRUE
library(caret)
library(gbm)
set.seed(0)
```

###2. Download training and test data


```{r}
trainingDataUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testDataUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

trainingDataFile <- "pml-training.csv"
testDataFile <- "pml-testing.csv"

if (!file.exists(trainingDataFile)) {
    download.file(trainingDataUrl, trainingDataFile, method="curl", quiet = T)
}

if (!file.exists(testDataFile)) {
    download.file(testDataUrl, testDataFile, method="curl", quiet = T)
}

inputData <- read.csv(trainingDataFile, sep = ",", quote = "\"")
testData <- read.csv(testDataFile, sep = ",", quote = "\"")
```

###3. Data cleaning

In order to reduce number of predictors and reduce noise, we remove some unimportant columns.

Before cleaning the dimensions of the training set are the following:

```{r}
dim(inputData)
```

Firstly, remove columns with "NA" values

```{r}
inputData <- inputData[, colSums(is.na(inputData)) == 0]
testData <- testData[, colSums(is.na(testData)) == 0]
```

Secondly, remove zero covariates
```{r}
nzvar <- nearZeroVar(inputData, saveMetrics = T)
inputData <- inputData[, !(nzvar$nzv)]

nzvar <- nearZeroVar(testData, saveMetrics = T)
testData <- testData[, !(nzvar$nzv)]
```

Lastly, remove columns which do not affect analyses
```{r}
remove <-  c("X", "user_name", "raw_timestamp_part_1", "raw_timestamp_part_2", "cvtd_timestamp")
inputData <- inputData[,!(names(inputData) %in% remove)] 
testData <- testData[,!(names(testData) %in% remove)]
```

So we can see that after cleaning the number of columns is reduced:

```{r}
dim(inputData)
```

###4. Data slicing

Create a data set that is 60% is allocated to the training set, and 40% is allocated to the testing set.

```{r}
inTrain <- createDataPartition(y = inputData$classe, p = 0.6, list = FALSE) 
training <-inputData[inTrain,] 
testing <- inputData[-inTrain,]
```

###5. Model creation

To predict the classe outcome, we use the boosting method, because it is a widely used and highly accurate method and provide better accuracy with less trees. It also allows to do  cross validation.
For this model we use 3-fold cross validation.

```{r}
myTuneGrid <- expand.grid(n.trees=seq(1,200,10), interaction.depth = 2:7, shrinkage = 0.1, n.minobsinnode = 10)
fitControl <- trainControl(method = "repeatedcv", number = 3, repeats = 1, verboseIter = FALSE, returnResamp = "all")
modFit <- train(classe ~ ., method = "gbm", data = training, trControl = fitControl, tuneGrid = myTuneGrid, verbose = FALSE)
````

#Results

After applying the model to the testing data set

```{r}
predictions <- predict(modFit, testing)
confusionMatrix(predictions, testing$classe)
accuracy <- postResample(predictions, testing$classe)
accuracy
ose <- 1 - as.numeric(confusionMatrix(testing$classe, predictions)$overall[1])
ose
```

we can see that the accuracy of the model is `r round(accuracy[1] * 100, digits = 2)`% and we expect the out of sample error to be `r round(ose * 100, digits = 2)`%.

### Test data set prediction

After applying our machine learning algorithm to the 20 test cases available in the test data above we can get predicted values:

```{r}
res <- predict(modFit, testData[, -length(names(testData))])
res
```

#Conclusion
With obtained values accuracy of `r round(accuracy[1] * 100, digits = 2)`% and the out of sample error of `r round(ose * 100, digits = 2)`% we can conclude that developed model provides quite high level accuracy.
